---
title: "Data Mining Project 3"
author: "Olivia Hofmann, Michael Perkins"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, warning = FALSE, message = FALSE)

# Set graphics device to avoid Quartz issues on macOS
options(device = "png")
```

## Introduction

The COVID-19 pandemic highlighted the importance of preparedness for infectious disease outbreaks. Anticipating which counties are at higher risk can enable early interventions, potentially saving lives and mitigating economic impacts. This project aims to classify U.S. counties into **high**, **medium**, or **low** risk categories for future pandemics based on historical COVID-19 data and other socioeconomic factors.

Following the CRISP-DM framework, we focus on:

 - Data Preparation: Defining classes, cleaning data, feature engineering, and handling missing values.
 - Modeling: Training and tuning multiple classification models.
 - Evaluation: Assessing model performance and their utility to stakeholders.
 - Deployment: Discussing practical application and maintenance of the models.
---

## 1. Data Preparoation

```{r load-libraries, cache=FALSE, message=FALSE, warning=FALSE}
# Load required libraries
library(tidyverse)
library(lubridate)
library(caret)
library(rpart)  # Decision Tree
library(randomForest)  # Random Forest
library(e1071)  # SVM
library(DMwR2)
library(PRROC)
library(ggplot2)
library(gridExtra)
library(pROC)
library(xgboost)
library(class)
library(nnet)
library(iml)
```

### Define Classes

- The classes are based on confirmed COVID-19 cases per 10,000 population per week:
  - **High Risk:** $\geq$ 50 cases.
  - **Medium Risk:** 10–49 cases.
  - **Low Risk:** < 10 cases.
- These thresholds were chosen based on observed patterns in case severity and the need to trigger timely interventions.

```{r load-data, cache=FALSE, , message=FALSE, warning=FALSE}
# Load mobility data
final_merged_dataset <- read_csv("data/final_merged_dataset.csv")

# Convert data from long to wide format
final_merged_dataset <- final_merged_dataset %>%
  pivot_wider(names_from = metric, values_from = count)

```

### Data Preparation Steps

1. **Merge and Clean Data**: Ensure a single dataset with a class attribute.
2. **Select Predictive Features**: Extract features with potential predictive power.
3. **Handle Missing Data**: Use imputation or remove incomplete rows for models that cannot handle missing data.

```{r process-data}
# Define risk levels
final_data <- final_merged_dataset %>%
  mutate(risk_level = case_when(
    confirmed_cases >= 50 ~ "high",
    confirmed_cases >= 10 ~ "medium",
    TRUE ~ "low"
  )) %>%
  mutate(risk_level = factor(risk_level, levels = c("low", "medium", "high")))

# Select relevant features
classification_data <- final_data %>%
  select(retail_and_recreation_percent_change_from_baseline,
         grocery_and_pharmacy_percent_change_from_baseline,
         workplaces_percent_change_from_baseline,
         PC1, PC2, week, risk_level) %>%
  drop_na()

# Split data into training and testing
set.seed(123)
training_index <- sample(1:nrow(classification_data), 0.7 * nrow(classification_data))
training_data <- classification_data[training_index, ]
testing_data <- classification_data[-training_index, ]

# Balance training data
min_class_size <- min(table(training_data$risk_level))
balanced_training_data <- training_data %>%
  group_by(risk_level) %>%
  sample_n(min_class_size) %>%
  ungroup()
```

---

## 2. Modeling

### Model 1: Decision Tree

- **Advantages**: Simple and interpretable.

```{r model-dt}
set.seed(123)
dt_model <- rpart(risk_level ~ ., data = balanced_training_data, method = "class",
                  control = rpart.control(cp = 0.05, maxdepth = 3))
```

### Model 2: Random Forest

- **Advantages**: Handles large datasets and captures feature interactions.

```{r model-rf}
set.seed(123)
rf_model <- randomForest(risk_level ~ ., data = balanced_training_data, ntree = 100, mtry = 2)
```

### Model 3: Support Vector Machine (SVM)

- **Advantages**: Effective for high-dimensional spaces.

```{r model-svm}
set.seed(123)
subsample_index <- sample(1:nrow(balanced_training_data), 0.01 * nrow(balanced_training_data))
subsample_data <- balanced_training_data[subsample_index, ]

svm_model <- svm(risk_level ~ ., data = subsample_data, cost = 0.1, gamma = 0.01, kernel = 'linear')
```

### Model 4: Gradient Boosting

- **Advantages**: 

```{r model-gb}
# Prepare training features and labels
x_train <- model.matrix(risk_level ~ . - 1, data = balanced_training_data)  # Remove intercept and non-numeric
y_train <- as.numeric(balanced_training_data$risk_level) - 1  # Convert factor to 0-indexed numeric

# Prepare testing features and labels
x_test <- model.matrix(risk_level ~ . - 1, data = testing_data)
y_test <- as.numeric(testing_data$risk_level) - 1

# Train the Gradient Boosting model
set.seed(123)
xgb_model <- xgboost(
  data = x_train,
  label = y_train,
  objective = "multi:softprob",  # Multiclass classification
  num_class = length(unique(balanced_training_data$risk_level)),  # Number of classes
  nrounds = 100,  # Number of boosting rounds
  eta = 0.1,      # Learning rate
  max_depth = 3,  # Tree depth
  verbose = 0     # Suppress training logs
)
```

### Model 5: Logistic Regression

- **Advantages**: 

```{r model-lr}
# Train a multinomial logistic regression model
logistic_model <- multinom(risk_level ~ ., data = balanced_training_data)
```

---

## 3. Evaluation

```{r evaluation functions, echo = FALSE}
# Enhanced function to calculate metrics and plot a bar chart
calculate_metrics <- function(conf_matrix, title = "Metrics Bar Chart") {
  # Calculate precision, recall, and F1-score
  precision <- diag(conf_matrix) / colSums(conf_matrix)  # TP / (TP + FP)
  recall <- diag(conf_matrix) / rowSums(conf_matrix)    # TP / (TP + FN)
  f1 <- 2 * (precision * recall) / (precision + recall) # F1 = 2 * (Precision * Recall) / (Precision + Recall)
  
  # Combine metrics into a list
  metrics <- list(
    Precision = precision,
    Recall = recall,
    F1 = f1
  )
  
  # Create a data frame for plotting
  metrics_df <- data.frame(
    Class = colnames(conf_matrix),
    Precision = precision,
    Recall = recall,
    F1 = f1
  )
  
  # Reshape data for ggplot
  metrics_long <- metrics_df %>%
    pivot_longer(cols = c("Precision", "Recall", "F1"), names_to = "Metric", values_to = "Value")
  
  # Generate a bar chart
  library(ggplot2)
  plot <- ggplot(metrics_long, aes(x = Class, y = Value, fill = Metric)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(title = title, x = "Class", y = "Score") +
    theme_minimal() +
    scale_fill_brewer(palette = "Set2")
  
  # Print the plot
  print(plot)
}

# Function to plot heatmap
plot_confusion_matrix <- function(conf_matrix, title) {
  # Convert the confusion matrix to a data frame
  data <- as.data.frame(as.table(conf_matrix))
  
  # Create a heatmap using ggplot2
  ggplot(data, aes(x = Predicted, y = Actual, fill = Freq)) +
    geom_tile() +
    geom_text(aes(label = Freq), color = "white", size = 4) +
    scale_fill_gradient(low = "blue", high = "red") +
    labs(title = title, x = "Predicted", y = "Actual") +
    theme_minimal()
}

# Function to calculate One-vs-All ROC curves
generate_roc_curves <- function(predictions, true_labels, model_name, classes) {
  library(pROC)
  
  roc_curves <- list()
  
  # Generate ROC for each class
  for (class in classes) {
    roc_curves[[paste0(model_name, " (", class, ")")]] <- roc(
      as.numeric(true_labels == class),
      as.numeric(predictions == class)
    )
  }
  
  return(roc_curves)
}

# Function to compute and plot PR curves
plot_pr_curve <- function(predictions, true_labels, model_name, class_label) {
  # Calculate PR curve
  pr <- pr.curve(
    scores.class0 = as.numeric(predictions == class_label),  # Predicted positive scores
    scores.class1 = as.numeric(true_labels == class_label),  # True positive labels
    curve = TRUE
  )
  
  # Plot the PR curve
  plot(pr, main = paste("Precision-Recall Curve:", model_name, "-", class_label))
  
  # Return PR curve object for further use
  return(pr)
}

# Function to compute misclassification for a model
compute_misclassification <- function(predictions, true_labels, model_name) {
  misclassified_data <- testing_data %>%
    mutate(
      Predicted = predictions,
      Misclassified = ifelse(Predicted != true_labels, "Yes", "No")
    ) %>%
    mutate(Model = model_name)  # Add model name for grouping

  return(misclassified_data)
}

# Function to save models.
save_model <- function(model, file_name) {
  saveRDS(model, file = file_name)
  cat("Model saved to:", file_name, "\n")
}

# Function to load models.
load_model <- function(file_name) {
  model <- readRDS(file_name)
  cat("Model loaded from:", file_name, "\n")
  return(model)
}
```

### 1. Confusion Matrix Heatmaps

```{r confusion-matrix-heatmap, echo=FALSE}
# Decision Tree
predictions_dt <- predict(dt_model, testing_data, type = "class")
dt_conf_matrix <- table(Predicted = predictions_dt, Actual = testing_data$risk_level)

# Random Forest
predictions_rf <- predict(rf_model, testing_data, type = "response")
rf_conf_matrix <- table(Predicted = predictions_rf, Actual = testing_data$risk_level)

# SVM
predictions_svm <- predict(svm_model, testing_data)
svm_conf_matrix <- table(Predicted = predictions_svm, Actual = testing_data$risk_level)

# Gradient Boosting
xgb_predictions <- predict(xgb_model, x_test)
xgb_probabilities <- matrix(xgb_predictions, ncol = length(levels(balanced_training_data$risk_level)), byrow = TRUE)
colnames(xgb_probabilities) <- levels(balanced_training_data$risk_level)
xgb_class <- apply(xgb_probabilities, 1, function(row) {
  colnames(xgb_probabilities)[which.max(row)]
})
xgb_conf_matrix <- table(Predicted = factor(xgb_class, levels = levels(balanced_training_data$risk_level)),
                         Actual = testing_data$risk_level)

# Logistic Regression
logistic_predictions <- predict(logistic_model, newdata = testing_data)
logistic_conf_matrix <- table(Predicted = logistic_predictions, Actual = testing_data$risk_level)


# Improved heatmap function with conditional legend
improved_heatmap <- function(conf_matrix, title, show_legend = TRUE) {
  data <- as.data.frame(as.table(conf_matrix))
  
  ggplot(data, aes(x = Predicted, y = Actual, fill = Freq)) +
    geom_tile(color = "white") +
    geom_text(aes(label = Freq), color = "black", size = 3) +
    scale_fill_gradient2(low = "lightblue", mid = "yellow", high = "red", midpoint = max(data$Freq) / 2) +
    labs(title = title, x = "Predicted", y = "Actual") +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 10),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = ifelse(show_legend, "right", "none")
    )
}

# Generate heatmaps with legends removed except for the first plot
heatmap_dt <- improved_heatmap(dt_conf_matrix, "Decision Tree Confusion Matrix", show_legend = FALSE)
heatmap_rf <- improved_heatmap(rf_conf_matrix, "Random Forest Confusion Matrix", show_legend = FALSE)
heatmap_svm <- improved_heatmap(svm_conf_matrix, "SVM Confusion Matrix", show_legend = FALSE)
heatmap_xgb <- improved_heatmap(xgb_conf_matrix, "Gradient Boosting Confusion Matrix", show_legend = FALSE)
heatmap_logistic <- improved_heatmap(logistic_conf_matrix, "Logistic Regression Confusion Matrix", show_legend = TRUE)

# Combine into a grid
grid.arrange(heatmap_dt, heatmap_rf, heatmap_svm, heatmap_xgb, heatmap_logistic, ncol = 2)
 
```

ALT

```{r confusion-matrix-objects, echo=FALSE}
# Generate predictions and confusion matrices for selected models
models <- list(
  "Decision Tree" = dt_model,
  "Random Forest" = rf_model,
  "Logistic Regression" = logistic_model
)

conf_matrices <- list()

for (model_name in names(models)) {
  model <- models[[model_name]]
  if (model_name == "Logistic Regression") {
    predictions <- predict(model, newdata = testing_data)
  } else {
    predictions <- predict(model, testing_data, type = "class")
  }
  conf_matrix <- table(Predicted = predictions, Actual = testing_data$risk_level)
  conf_matrices[[model_name]] <- conf_matrix
}


```

---
---

### 2. Precision, Recall and F1-score
```{r precision-recall-F1score, echo =FALSE}
# Decision Tree
dt_metrics <- calculate_metrics(as.matrix(dt_conf_matrix), title = "Decision Tree Metrics")

# Random Forest
rf_metrics <- calculate_metrics(as.matrix(rf_conf_matrix), title = "Random Forest Metrics")

# SVM
svm_metrics <- calculate_metrics(as.matrix(svm_conf_matrix), title = "SVM Metrics")

# Gradient Boosting
xgb_metrics <- calculate_metrics(as.matrix(xgb_conf_matrix), title = "Gradient Boosting Metrics")

# Logistic Regression
logistic_metrics <- calculate_metrics(as.matrix(logistic_conf_matrix), title = "Logistic Regression Metrics")
```

ALT 

```{r precision-recall-F1score-alt, echo =FALSE}
library(knitr)

calculate_metrics_table <- function(conf_matrix, model_name) {
  # Calculate precision, recall, and F1-score
  precision <- diag(conf_matrix) / colSums(conf_matrix)  # TP / (TP + FP)
  recall <- diag(conf_matrix) / rowSums(conf_matrix)     # TP / (TP + FN)
  f1 <- 2 * (precision * recall) / (precision + recall)  # F1 = 2 * (Precision * Recall) / (Precision + Recall)
  
  metrics_df <- data.frame(
    Model = model_name,
    Class = rownames(conf_matrix),
    Precision = round(precision, 2),
    Recall = round(recall, 2),
    F1_Score = round(f1, 2)
  )
  return(metrics_df)
}

# Collect metrics for each model
metrics_list <- list()
for (model_name in names(conf_matrices)) {
  conf_matrix <- conf_matrices[[model_name]]
  metrics_df <- calculate_metrics_table(conf_matrix, model_name)
  metrics_list[[model_name]] <- metrics_df
}

# Combine metrics into a single table
all_metrics <- do.call(rbind, metrics_list)

# Display the metrics table
kable(all_metrics, caption = "Precision, Recall, and F1-Score for Each Model")


```

### 2. ROC Curves

```{r roc-curves, echo=FALSE, warning=FALSE, message=FALSE}
classes <- c("low", "medium", "high")

# Decision Tree
roc_curves_dt <- generate_roc_curves(predictions_dt, testing_data$risk_level, "Decision Tree", classes)

# Random Forest
roc_curves_rf <- generate_roc_curves(predictions_rf, testing_data$risk_level, "Random Forest", classes)

# SVM
roc_curves_svm <- generate_roc_curves(predictions_svm, testing_data$risk_level, "SVM", classes)

# Gradient Boosting (requires probabilities)
xgb_class_prob <- predict(xgb_model, x_test) # Predict probabilities
xgb_class_prob_matrix <- matrix(xgb_class_prob, ncol = length(unique(y_train)), byrow = TRUE) # Reshape into a matrix (rows = instances, columns = classes)
colnames(xgb_class_prob_matrix) <- levels(balanced_training_data$risk_level) # Assign column names based on class levels
roc_curves_xgb <- list(
  "Gradient Boosting (low)" = roc(as.numeric(testing_data$risk_level == "low"), xgb_class_prob_matrix[, "low"]),
  "Gradient Boosting (medium)" = roc(as.numeric(testing_data$risk_level == "medium"), xgb_class_prob_matrix[, "medium"]),
  "Gradient Boosting (high)" = roc(as.numeric(testing_data$risk_level == "high"), xgb_class_prob_matrix[, "high"])
)

# Logistic Regression
roc_curves_lr <- generate_roc_curves(logistic_predictions, testing_data$risk_level, "Logistic Regression", classes)

# Combine Gradient Boosting ROC curves with others
all_roc_curves <- c(roc_curves_dt, roc_curves_rf, roc_curves_svm, roc_curves_xgb, roc_curves_lr)

# Combine all ROC curves into a single data frame for ggplot
all_roc_curves_df <- bind_rows(
  lapply(names(all_roc_curves), function(name) {
    data.frame(
      specificity = 1 - all_roc_curves[[name]]$specificities,
      sensitivity = all_roc_curves[[name]]$sensitivities,
      name = name
    )
  })
)

facet_labels <- c(low = "Low", medium = "Medium", high = "High")

# Plot with facets for classes
ggplot(all_roc_curves_df, aes(x = specificity, y = sensitivity, color = gsub(" \\(.*\\)", "", name), linetype = substr(name, regexpr("\\(", name) + 1, regexpr("\\)", name) - 1))) +
  geom_line(size = 1) +
  labs(title = "One-vs-All ROC Curves for All Models", x = "1 - Specificity", y = "Sensitivity", color = "Model", linetype = "Class") +
  theme_minimal(base_size = 14) + # Adjust font size
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    legend.text = element_text(size = 10) # Control text size
  ) +
  facet_wrap(~ substr(name, regexpr("\\(", name) + 1, regexpr("\\)", name) - 1), 
             scales = "fixed", 
             labeller = as_labeller(facet_labels))


```



```{r roc-curves-table, echo=FALSE, warning=FALSE, message=FALSE}
library(pROC)

# Function to calculate AUC for each model and class
calculate_auc <- function(predictions, true_labels, model_name, classes) {
  auc_list <- list()
  for (class in classes) {
    roc_obj <- roc(as.numeric(true_labels == class), as.numeric(predictions == class))
    auc_value <- auc(roc_obj)
    auc_list[[class]] <- auc_value
  }
  auc_df <- data.frame(
    Model = model_name,
    Class = classes,
    AUC = unlist(auc_list)
  )
  return(auc_df)
}

# Calculate AUC values for each model
classes <- c("low", "medium", "high")
auc_metrics <- list()

for (model_name in names(models)) {
  model <- models[[model_name]]
  if (model_name == "Logistic Regression") {
    predictions <- predict(model, newdata = testing_data)
  } else {
    predictions <- predict(model, testing_data, type = "class")
  }
  auc_df <- calculate_auc(predictions, testing_data$risk_level, model_name, classes)
  auc_metrics[[model_name]] <- auc_df
}

# Combine AUC metrics into a single table
all_auc_metrics <- do.call(rbind, auc_metrics)

# Display the AUC table
kable(all_auc_metrics, caption = "AUC Values for Each Model and Class")

```
---

### 3. Precision-Recall Curves

```{r precision-recall, echo=FALSE}
classes <- c("low", "medium", "high")  # Class labels

# Decision Tree
pr_curves_dt <- lapply(classes, function(class) plot_pr_curve(predictions_dt, testing_data$risk_level, "Decision Tree", class))

# Random Forest
pr_curves_rf <- lapply(classes, function(class) plot_pr_curve(predictions_rf, testing_data$risk_level, "Random Forest", class))

# SVM
pr_curves_svm <- lapply(classes, function(class) plot_pr_curve(predictions_svm, testing_data$risk_level, "SVM", class))

# Gradient Boosting (using probabilities)
xgb_class_prob_matrix <- matrix(predict(xgb_model, x_test), ncol = length(classes), byrow = TRUE)
colnames(xgb_class_prob_matrix) <- classes
pr_curves_xgb <- lapply(classes, function(class) {
  plot_pr_curve(
    predictions = xgb_class_prob_matrix[, class] > 0.5,  # Threshold probabilities at 0.5
    true_labels = testing_data$risk_level,
    model_name = "Gradient Boosting",
    class_label = class
  )
})

# Logistic Regression
pr_curves_lr <- lapply(classes, function(class) plot_pr_curve(logistic_predictions, testing_data$risk_level, "Logistic Regression", class))
```

---

### 4. Feature Importance

```{r feature-importance, echo=FALSE}
# Decision Tree

# Random Forest
rf_importance <- as.data.frame(importance(rf_model))
rf_importance$Feature <- rownames(rf_importance)
ggplot(rf_importance, aes(x = reorder(Feature, MeanDecreaseGini), y = MeanDecreaseGini)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Feature Importance (Random Forest)", x = "Features", y = "Importance") +
  theme_minimal()

# SVM

# Gradient Boosting
xgb_importance <- xgb.importance(model = xgb_model, feature_names = colnames(x_train))
xgb_importance_df <- as.data.frame(xgb_importance) # Convert to Data Frame
ggplot(xgb_importance_df, aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_bar(stat = "identity", fill = "darkorange") +
  coord_flip() +
  labs(title = "Feature Importance (Gradient Boosting)", x = "Features", y = "Gain") +
  theme_minimal()

# Logistic Regression

```

---

### 5. Misclassification Analysis

```{r misclassification-analysis, echo=FALSE}
# Decision Tree
misclassified_dt <- compute_misclassification(predictions_dt, testing_data$risk_level, "Decision Tree")

# Random Forest
misclassified_rf <- compute_misclassification(predictions_rf, testing_data$risk_level, "Random Forest")

# SVM
misclassified_svm <- compute_misclassification(predictions_svm, testing_data$risk_level, "SVM")

# Gradient Boosting
xgb_probabilities <- matrix(xgb_predictions, ncol = length(levels(testing_data$risk_level)), byrow = TRUE)
xgb_class_predictions <- apply(xgb_probabilities, 1, function(row) {
  levels(testing_data$risk_level)[which.max(row)]
})
misclassified_xgb <- compute_misclassification(
  factor(xgb_class_predictions, levels = levels(testing_data$risk_level)),
  testing_data$risk_level,
  "Gradient Boosting"
)

# Logistic Regression
misclassified_logistic <- compute_misclassification(logistic_predictions, testing_data$risk_level, "Logistic Regression")

# Combine misclassification data
misclassified_combined <- bind_rows(
  misclassified_dt,
  misclassified_rf,
  misclassified_svm,
  misclassified_xgb,
  misclassified_logistic
)

# Visualize misclassifications
ggplot(misclassified_combined, aes(x = risk_level, fill = Misclassified)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ Model) +
  labs(title = "Misclassification Analysis by Model and Risk Level",
       x = "Risk Level",
       y = "Count") +
  theme_minimal()

```

---

### Conclusion

These visuals provide insights into the models' performance and help stakeholders understand the trade-offs between accuracy, precision, and recall for each classification method. They also highlight areas for improvement, such as addressing misclassifications.

---

## 4. Deployment

- **Practical Use**: The model can guide early interventions (e.g., mask mandates, closures).
- **Update Frequency**: Weekly updates based on new data.
- **Integration**: Stakeholders can incorporate model predictions into decision-making frameworks.

```{r deployment}
# Save all models
save_model(dt_model, "dt_model_balanced.rds")
save_model(rf_model, "rf_model_balanced.rds")
save_model(svm_model, "svm_model_balanced.rds")
save_model(xgb_model, "xgb_model_balanced.rds")
save_model(logistic_model, "logistic_model_balanced.rds")

# Load all models
loaded_dt_model <- load_model("dt_model_balanced.rds")
loaded_rf_model <- load_model("rf_model_balanced.rds")
loaded_svm_model <- load_model("svm_model_balanced.rds")
loaded_xgb_model <- load_model("xgb_model_balanced.rds")
loaded_logistic_model <- load_model("logistic_model_balanced.rds")
```

---

## Appendix

- **Team Contributions**:
  - Olivia Hofmann: Lead on data preparation and feature engineering.
  - Michael Perkins: Lead on modeling and evaluation.

- **Graduate Work**:
  - Additional models: Gradient Boosting and k-Nearest Neighbors (to be implemented).
